The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-18 manuscript/manuscript.Rnw
move the discussion of the go analysis for core
remove extra citations to get to 45

Even though the placenta's primary function is conserved among
mammals, it varies widely in terms of the shape of attachment, degree
of intimacy between fetal tissues and maternal tissues and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often with specific
aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

Previous studies of some species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have provided some
preliminary answers to these questions. They revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
in \human and \mouse, \acp{PAG} in \cow, genes related to the
thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
  africana} (in this work), and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria (or in the
absence of proteinuria, thrombocytopenia, impaired liver function,
renal insufficiency, pulmonary edema, or cerebral or visual
disturbances) %
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. %
Although there are speculative cases of eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

>>>>

Even though the placenta's primary function is conserved among
mammals, it varies widely in terms of the shape of attachment, degree
of intimacy between fetal tissues and maternal tissues and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often with specific
aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

Previous studies of some species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have provided some
preliminary answers to these questions. They revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
in \human and \mouse, \acp{PAG} in \cow, genes related to the
thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
  africana} (in this work), and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}). Supplemental
\cref{tab:core_placenta_pathway_analysis} shows the pathways with
significant and suggestive enrichment in the core placenta
transcriptome.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria (or in the
absence of proteinuria, thrombocytopenia, impaired liver function,
renal insufficiency, pulmonary edema, or cerebral or visual
disturbances) %
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. %
Although there are speculative cases of eclampsia in great apes, 
%\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\citeC{arter.ea2011:Evolutionofinvasiveplacentation}.
